Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-03-18 manuscript/manuscript.Rmd
updated the outline and readme file.

# Introduction

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following four measures as integral components of the Triple P framework: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to demonstrate the necessity and potentials of this framework.

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020]. The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

# The ethology and ethnography of stone toolmaking experiments

@@

Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].

@@

>>>>

\vspace{12pt}

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. The Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following four measures as integral components of the Triple P framework: 1) normalizing the ethological and ethnographic data collection in experimental projects ; 2) encouraging the involvements of avocational as well as novice participants; 3) boosting the collaboration across labs in a global scale; 4) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potentials of this framework.

# What good is actualistic experiment?

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016; @lin2018; @rezek2020; @outram2008]. The controlled experiments conducted on stone artifacts [@li2022], partcularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights regarding our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of human behaviors due to the messiness of the latter. In the past decades, actualistic experiments becomes more common [@liu2022]. Is RCT the golden standard of knowledge [@cartwright2007]

@@

# The ethology and ethnography of stone toolmaking

@@

Experimental archaeology as a scientific method is rooted in the individualistic reverse engineering in the 19th century instead of inter-generation transmission of knapping knowledge that spans several million years [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61].

@@

# Acknowledgements

This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" (C. L.).
